Assessing the Impact of MDE and Code Generation on the Sustainability of Scientific Computing Software: A Research Proposal

### 2019-09-25 OverallResearchProposal/ResearchProposal.tex
Reorg folder structure to put research proposal in its own folder

The scope of SCS software for this proposal includes long-lived software, as
implied by the sustainability objective.  To gain the full benefits of
generating documentation, the scope also includes software that interests
multiple stakeholders, with different interests and backgrounds.  Finally, the
scope of this proposal emphasizes safety related software, such as software for
nuclear safety analysis, medical imaging and computational medicine.  Safety
related software is prioritized because documentation is often expected as part
of a certification exercise.  Although this study is not specifically restricted
to open-source software, for practical reasons (especially the lack of access to
the code), commercial software will not be strongly emphasized.

@@

\item [\textbf{Productivity}] [Fill in]

\item [\textbf{Sustainability}] A combination of other qualities, likely
  maintainability and productivity.  Reference authorless paper found on-line.

@@

MEng students will be asked to assess the state of practice in each domain.
During the last state of the practice exercise each student measured two
domains, but with the higher demands from the new measurement protocol, we will
likely switch to one domain per student.

In the previous project, we measured 30 software projects for each domain.  With
the increased scrutiny required in this re-boot, we will not likely be able to
measure the details on this high a number of software projects.  We may still
start with 30 software examples in each domain, but then use some criteria to
create a shorter sub-list for detailed measurement.  The details of how to
proceed here would be determined and documented in the measurement protocol.

@@

\begin{itemize}
\item Characterization of the functionality provided by the software in the
  domain.  What services does each member of the domain provide?  Ideally this
  information would be summarized in a commonality analysis document.  The
  document will summarize the domain software via its commonalities,
  variabilities and parameters of variation.
\item Usability testing of each software in the domain.  Specific metrics for
  assessing usability still need to be researched, but measures that involve the
  time to complete a task and the users overall impression will be considered.
\item Collect empirical software engineering related data, such as the number of
  files, number of lines of code, cyclomatic complexity, number of open issues,
  etc.  As a starting point for tool support, HubListener
  (\url{https://github.com/pjmc-oliveira/HubListener}) could be used.
\item Collect at least one empirical measure of the quality of the documentation
  - the number of lines of documentation.  We should investigate to see if this
  is a metric that has previously been collected.  So many of the existing
  metrics are completely focused on the code, or possibly on issue creation and
  management; it seems that little work has been done to assess the
  documentation quality.
\item One of the guidelines for successfully on-boarding new developers is to
  ``Keep knowledge up to date and findable'' \citep{ShollerEtAl2019}.  The
  measurements selected should be able to assist with judging whether a project
  succeeds on this front.
\end{itemize}

The above data will be combined to rank the software in the domain.  The
specifics are yet to be determined, but in the past the Analytic Hierarchy
Process (AHP) \citep{Saaty1980} proved helpful in this.  In the past the AHP
results were calculated by a grade out of 10 that each student evaluator gave to
each quality for each measured product.  A better approach may be to have a
group of people familiar with the software do a pairwise comparison between each
product to come up with a number.  The number would be based on the data
collected, but not in a numerical way.  The data collected would sever the same
role as a decision support system.  It would provided insight, but human beings
would do the final pairwise comparisons.

@@

As a next step for this work, the procedure for measuring software qualities in
SCS will be brainstormed, discussed and written up.  The procedure will be
applied to each domain with SCS that is assessed.  The write-up on the procedure
will be published.  Ideally at a relevant conference, but if that is not a
feasible option, it will be hosted on the arxiv.

@@

\begin{itemize}
\item Availability of a domain expert that is willing to assist, and who has at
  least a few graduate students that can assist with the measurement tasks.
\item A domain where end-user developed software is common
\item A domain where open-source software is available.  Commercial products are
  not as effective for measurement, since we cannot generally gain access to the
  source code, developer documentation, or issue tracking system.
\item A domain where scientific computing is used.  The selected software should
  fit in the larger domain of SCS.
\end{itemize}

@@

The general idea that MDE will be helpful is appealing, but to study the impact
of MDE, we need to know how it will actually be used.  How exactly would MDE be
fit into the scientific software development process?  The new process should
encourage an early intensity of thought and effort, but at the same time, it
should not impose waterfall development, since waterfall is an unrealistic
expectation for a problem that is not yet well understood.  MDE supports
frequent change and incremental development, but where should developers start
and how should the work proceed?  How does one create an almost empty project in
MDE and then borrow from existing knowledge and create new knowledge?  To create
a specific MDE process for SCS, the Drasil tools will be utilized.  A good
starting project could be
\href{https://jacquescarette.github.io/Drasil/examples/Projectile/srs/Projectile_SRS.html}
{Projectile}, which models simple projectile motion.

@@

@@

We will identify multiple case studies, with potential projects coming from the
domains listed for the ``state of the practice'' exercise
(Section~\ref{SecStateOfThePractice}), like computational medicine, medical
imaging or quantum chemistry software.  We will first identify the partners and
look for feasible projects that can be developed using an MDE approach.  We will
use the ideal process identified in Section~\ref{SecMDEProcess}.  Compare to Some additional thought, and resources, are necessary
here.  If we could do a project where the scientist, or a student under their
supervision, is a true partner in the development, we might get more convincing
anecdotal evidence.  Pre and post interviews.  Measurement of quality following the
techniques developed in the state of the practice exercise.

List of projects above could also be used for case studies.

Observe - how answer questions when problems come up.  Should have so
pre and post interviews.

\subsubsection{Development of Small Physics
  Examples} \label{SecSmallPhysExample}

describe a different experiment.  conference multiple develop physics code.
Like projectile.  Survey before and after.  Quantify success of efforts and time required.

@@

>>>>

The scope of SCS software for this proposal includes software intended for a
long life (10+ years), as implied by the sustainability objective.  The effort
involved in MDE does not provide enough of a payoff for short-term projects.  To
gain the full benefits of generating documentation, the scope also includes
software that interests multiple stakeholders, with different interests and
backgrounds.  Finally, the scope of this proposal emphasizes safety related
software, such as software for nuclear safety analysis, medical imaging and
computational medicine.  Safety related software is prioritized because
documentation is often expected as part of a certification exercise.  Although
this study is not specifically restricted to open-source software, for practical
reasons (especially the lack of access to the code), commercial software will
not be strongly emphasized.

Implicit in the scope is that we are interested in software for domains that are
well-understood.  This is necessary for knowledge capture and the associated
construction of models.  We can refine the notion of well-understood by using
the fact that MDE is simply a technology based realization of the older concept
of program families.  Success with an MDE approach requires
satisfaction of the same three hypotheses as success for program family
development \citep{Weiss1997}:

\begin{description}
\item [Redevelopement Hypothesis] -- most software development involved in
  producing an individual family members should be redevelopement because the
  family members have so much in common.

\item[Oracle Hypothesis] -- the changes that are likely to occur during the
  software's lifetime are predictable.

\item [Organizational Hypothesis] -- designers can organize the software and the
  development effort so that predicted changes can be made independently.
\end{description}

SCS can be shown to satisfy these three hypotheses \citep{SmithAndChen2004,
  SmithMcCutchanAndCao2007}; therefore, SCS is suitable for a program family, or
MDE software development approach.  Although our focus is on SCS, an MDE
approach could be applied to any domain that satisfies the above three
hypotheses.

@@

\item [\textbf{Productivity}] \wss{This needs to be filled in.  Productivity is
    an important quality for this proposal, since part of what MDE promises is
    increased productivity.}

\item [\textbf{Sustainability}] \wss{This needs to be filled in.  A search needs
    to be done to find a good definition.  A starting point might be the recent
    paper that defines sustainability as a combination of other qualities.}

@@

MEng students will be asked to assess the state of practice in each domain.
During the previous state of the practice exercise each student measured two
domains, but with the higher demands from the new measurement protocol, we will
likely switch to one domain per student.

In the previous project, we measured 30 software projects for each domain.  With
the increased scrutiny required in this re-boot, we will not likely be able to
measure the details on this high a number of software projects.  We may still
start with 30 software examples in each domain, but then use some criteria to
create a shorter sub-list for detailed measurement.  The details of how to
proceed here would be determined and documented in the measurement procedures.
We still would like to keep the number of measured software projects above 10.

@@

\begin{itemize}
\item Characterization of the functionality provided by the software in the
  domain.  What services does each member of the domain provide?  Ideally this
  information would be summarized in a commonality analysis document.  The
  document will summarize the domain software via its commonalities,
  variabilities and parameters of variation.
\item Usability testing of each software in the domain.  Specific metrics for
  assessing usability still need to be researched, but measures that involve the
  time to complete a task and the users overall impression will be considered.
\item Collect empirical software engineering related data, such as the number of
  files, number of lines of code, cyclomatic complexity, number of open issues,
  number of closed issues, etc.  As a starting point for tool support,
  HubListener (\url{https://github.com/pjmc-oliveira/HubListener}) could be
  used.
\item Collect at least one empirical measure of the quality of the documentation
  - the number of lines of documentation.  We should investigate to see if this
  is a metric that has previously been collected.  So many of the existing
  metrics are completely focused on the code, or possibly on issue creation and
  management; it seems that little work has been done to assess the
  documentation quality.
\item One of the guidelines for successfully on-boarding new developers is to
  ``Keep knowledge up to date and findable'' \citep{ShollerEtAl2019}.  The
  selected measurements should be able to assist with judging whether a project
  succeeds on this front.
\end{itemize}

To focus the revised measurement procedure, we will go through each of the
quality criteria listed in Section~\ref{SecSoftwareQuality}.  For each criteria
we will look at how it was previously measured by us, but also at the literature
and current best practices for its measurement.

The above data will be combined to rank the software in the domain.  The
specifics are yet to be determined, but in the past the Analytic Hierarchy
Process (AHP) \citep{Saaty1980} proved helpful in this.  In the past the AHP
results were calculated by a grade out of 10 from each student evaluator for
each quality for each measured product.  A better approach may be to have a
group of people familiar with the software do a pairwise comparison between each
product to come up with a number.  The number would be based on the data
collected, but not in a numerical way.  The data collected would serve the same
role as a decision support system.  It would provided insight, but human beings
would do the final pairwise comparisons, and thus the final ranking.

@@

As a next step for this work, the procedure for measuring software qualities in
SCS will be brainstormed, discussed and written up.  The procedure will be
applied to each domain within SCS that is assessed.  The write-up on the procedure
will be published.  Ideally at a relevant conference, but if that is not a
feasible option, it will be posted on the arxiv.

@@

\begin{itemize}
\item Availability of a domain expert that is willing to assist, and who has at
  least a few graduate students that can assist with the measurement tasks.
\item A domain where end-user developed software is common.
\item A domain where open-source software is available.  Commercial products are
  not as effective for measurement, since we cannot generally gain access to the
  source code, developer documentation, or issue tracking system.
\item A domain where at least 10 software projects are available for
  measurement.
\item A domain where scientific computing is used.  The selected software should
  fit in the larger domain of SCS.
\end{itemize}

@@

The general idea that MDE will be helpful is appealing, but to study the impact
of MDE, we need to know how it will actually be used.  How exactly would MDE be
fit into the scientific software development process?  The new ideal process
should encourage an early intensity of thought and effort, but at the same time,
it should not impose waterfall development, since waterfall is an unrealistic
expectation for a problem that is not yet well understood.  MDE supports
frequent change and incremental development, but where should developers start
and how should the work proceed?  How does one create an almost empty project in
MDE and then borrow from existing knowledge and create new knowledge?  To create
a specific MDE process for SCS, the Drasil tools will be utilized.  A good
starting project could be
\href{https://jacquescarette.github.io/Drasil/examples/Projectile/srs/Projectile_SRS.html}
{Projectile}, which models simple projectile motion.

@@

The ideal process using the Drasil tool will focus on the ideal version of
Drasil, since Drasil is still a work on progress.  The design of the ideal
process for the ideal Drasil, will help to guide future development of Drasil.

@@

We will identify multiple case studies, with potential projects coming from the
domains listed for the ``state of the practice'' exercise
(Section~\ref{SecStateOfThePractice}), like computational medicine, medical
imaging or quantum chemistry software.  We will first identify the partners and
look for feasible projects that can be developed using an MDE approach.  We will
use the ideal process identified in Section~\ref{SecMDEProcess}.  The respective
advantages and disadvantages of Drasil versus Epsilon will be compared.

To assess the impact of MDE, we will use techniques like pre and post
interviews, the measurements of quality used for the ``state of the practice''
exercise, ethnographic study,  etc.

\wss{More information is needed in this section.}

\subsubsection {Development of Small Physics Examples} \label{SecSmallPhysExample}

To learn more about MDE for SCS, and to assess the current and future state of
Drasil, a workshop could be conducted where the participants are all expected to
build a small physics application, together with full documentation and test
cases, in a one day, or possibly half day workshop.  The participants would be
surveyed before and after the development exercise, and observations would be
made throughout the exercise.

\wss{More detail is needed.}

@@

\wss{Schedules are incomplete}

\subsection{State of Practice Exercise}

\begin{itemize}
\item Fall 2019
  \begin{itemize}
  \item Review previous measurement template
  \item Update measurement process
  \item Meet with stakeholders
  \item Identify list of software to be measured
  \end{itemize}
\item Winter 2020
  \begin{itemize}
  \item Complete measurements
  \item Use AHP to rank software products for each quality, and for overall
    quality
  \end{itemize}
\item Summer 2020
  \begin{itemize}
  \item Write up results, MEng reports, papers
  \end{itemize}
\end{itemize}

\subsection{Impact of MDE on the Sustainability Timeline}

A high-level timeline of the research plan for assessing the impact of MDE on
sustainability (see Section~\ref{SecImpactMDE} ) follows:

\begin{itemize}
\item Fall 2019
  \begin{itemize}
  \item Write research protocol
  \end{itemize}
\item Winter 2020
  \begin{itemize}
  \item 
  \end{itemize}
\item Summer 2020
  \begin{itemize}
  \item 
  \end{itemize}
\item Fall 2020
\item Winter 2021
\item Summer 2021
\item Fall 2021
\end{itemize}
